Keep your Laziness in Check

### 2018-02-12 paper/main.tex
[minor] Add acknowledgement for Robert and José
[paper] Lazy/Strict Maps

\leo{The singular feels weird here. Is it just me?}
Purely functional queue is a very popular data structure in
Haskell. Programmers typically implement such queues with two
lists---a front list holding the queue elements in FIFO order, and a
back list holding the other elments in reversed order. When the front
list is empty, a dequeue operation requires a reversal of the back
list to refill the front list. These specific dequeue operations take
$O(n)$ time to finish, and are much slower than their imperative
counterparts. This occasional performance degradation is compounded by
persistence (where the $O(n)$ operation could be repeated many times),
and can be unacceptable in real-time systems.

@@

\leo{Example 2: Knapsack}

Laziness also shines in solutions to dynamic programming problems. Let
us consider a solution to the famous knapsack problem that levareges
laziness.

The inputs of the knapsack problem are two sequences of numbers $w_i$
and $v_i$ where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify
the weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.

@@

Using the lazy variant of \lk{Map} here is critical for the correctness
of \lk{solution}. If we used \lk{Data.Map.Strict} instead, then \lk{fromList}
would have eagerly evaluated the \lk{if} expression. The \lk{if} expression
contains value-level recursion to itself, and this would manifest as an infinite
loop at runtime. This is an incredibly subtle bug related to laziness.

>>>>

\leo{The singular feels weird here. Is it just me?}
As a first example, consider purely functional queues, a very
popular data structure in Haskell. Programmers typically implement
such queues with two lists---a front list holding the queue elements
in FIFO order, and a back list holding the other elments in reversed
order. When the front list is empty, a dequeue operation requires a
reversal of the back list to refill the front list. These specific
dequeue operations take $O(n)$ time to finish, and are much slower
than their imperative counterparts. This occasional performance
degradation is compounded by persistence (where the $O(n)$ operation
could be repeated many times), and can be unacceptable in real-time
systems.

@@

\leo{Example 2: Maps / Knapsack}

For the second example, we turn to a different data functional data
structure where both a strict and a lazy variant can be situationally
useful: maps. There are two reasonable laziness behaviors that
implement a map interface; we could implement a completely strict map,
or we could lazily evaluate the values stored as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing bottoms
to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}

Moreover, it is important to be able to distinguish whether such a
datatype follows to a desired laziness behavior. For instance, the
standard functional solution for dynamic programming problems
crucially uses laziness. In particular, consider the famous knapsack
problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.

@@

\leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
I thought the problem was that it computes everything when it doesn't need to}
Using the lazy variant of \lk{Map} here is critical for the correctness
of \lk{solution}. If we used \lk{Data.Map.Strict} instead, then \lk{fromList}
would have eagerly evaluated the \lk{if} expression. The \lk{if} expression
contains value-level recursion to itself, and this would manifest as an infinite
loop at runtime. This is an incredibly subtle bug related to laziness.
